The Precambrian paleogeography of Laurentia

### 2021-06-01 Manuscript/Laurentia_Paleogeo_Manuscript.tex
add ECMB pole to tables and Fig 1 map
further integrate ECMB results and make other small text edits

Growth of Laurentia in the Paleoproterozoic also occurred through accretionary orogenesis. In the northwest, this accretion occurred within the Wopmay orogen through ca. 1.88 Ga arc-continent collision that led to the accretion of the Hottah terrane (the Calderian orogeny) and the subsequent emplacement of the Great Bear magmatic zone from ca. 1.88 to 1.84 Ga \citep{Hildebrand2009a}. Coeval with the Trans-Hudson orogeny was accretionary orogenesis on the southern margin of the Superior province (the Penokean orogeny; \citealp{Schulz2007a}) and the southern margin of the North Atlantic province (the Makkovik-Ketilidian orogeny; \citealp{Kerr1996a}). The Penokean orogeny involved the accretion of a microcontinent block (the Marshfield terrane) and arc terranes to the southern margin of the west Superior province ca. 1.86 to 1.82 Ga (Fig. \ref{fig:Laurentia_map}; \citealp{Schulz2007a}). Firm evidence of the end of the Penokean orogeny comes from the ca. 1.78 Ga undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a}.

@@

Extensive rifting followed by thermal subsidence occurred along the southeastern to eastern Laurentia margin in the time leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean. On the southern part of the eastern margin, a record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig. \ref{fig:Laurentia_map}) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a, Whitmeyer2007a}. The ages of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this history such as in the rift development model of \citet{Burton2010a}. In this model, spatially-restricted rifting occurs ca. 760 to 680 Ma in the region of present-day North Carolina and Virginia. Rifting initiated in the region from present-day New York to Newfoundland ca. 620 to 580 Ma and by ca. 580 to 550 Ma rifting extended along the length of Laurentia's eastern margin. The last phase of this rifting has been interpreted to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane ca. 540 Ma \citep{Dickerson1998a, Martin2019a}. Rifting of the Laurentia margin north of the New York promontory was occuring by ca. 560 Ma based on the age of volcanics that are interlayered with sedimentary rocks interpreted to be synrift deposits \citep{Hodych2007a}.

As with other rifted margins, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting and departure of a major craton, as the record that lingers on the craton is similar. Recognizing this ambiguity, it has been proposed that rather than being associated with spatially-restricted or failed rifting, ca. 700 Ma extension was associated with breakup and separation of a conjugate continent and that there was subsequent rifting of smaller terranes off the eastern Laurentia margin ca. 600 Ma \citep{Chew2008a, Escayola2011a, Robert2020a}. In these models, the separation of a conjugate continent ca. 700 Ma (interpreted to be Amazonia) leads to the formation of an ocean termed the Puncoviscana Ocean by \citet{Escayola2011a} and the Paleo-Iapetus Ocean by \citet{Robert2020a}. The subsequent rifting of terranes off the Laurentia margin ca. 600 Ma would have resulted in the formation of the (Neo-)Iapetus Ocean and the record of thermal subsidence associated with passive margin development on Laurentia \citep{Escayola2011a, Robert2020a}.

@@

In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications (Fig. \ref{fig:Laurentia_poles} and Table 2). The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades to convey the confidence that the community has in these results \citep{Evans2021a}. While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland \citep{Brown2018a}. The `A' rating refers to poles that are judged to be of such high quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. The `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints. In addition, in this chapter I refer to select additional poles that were not given an `A' or `B' classification at the Nordic Workshops as not-rated (`NR'). These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a} in conjunction with the papers in which they were reported. Many additional poles in the Paleomagia database are valuable and should not be dismissed from being considered in paleogeographic reconstructions. However, there are ambiguities associated with many of the poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained, including its age. For example, there are many poles that have been developed from lithologies of the Trans-Hudson orogen (e.g. \citealp{Symons2005a}. However, these poles typically lack field tests, have poor control on paleohorizontal and are hypothesized to suffer from variable overprints as discussed in \cite{Raub2008a} and \cite{DAgrella-Filho2020a}. Given the preponderance of similar directions, three representative poles from the Trans-Hudson orogen are included in the Nordic Workshop compilation (Table 2) as `B' rated poles. However, they should be used with caution given that they suffer from the same deficiencies.

There are similar issues with the rich data sets associated with intrusive and metamorphic lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary. The ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and associated cooling within the Grenville orogen. Cooling ages of deeply exhumed lithologies are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions. As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating with the exception of the `B' rated pole from the ca. 1015 Ma Haliburton intrusions (Table 2). However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within paleogeographic reconstructions of Laurentia (although an alternative view of an allochthonous origin put forward by \citealp{Halls2015a} is discussed below). In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands for which the magnetic mineralogy and associated relative ages of remanence are relatively well-constrained are included (Table 2). An additional not-rated pole included in the present compilation is a more recently published pole for the ca. 1144 Ma Ontario lamprophyre dikes \citep{Piispa2018a} that strengthens the position of Laurentia at the time and coincides with the position of the poles from the ca. 1140 Ma Abitibi dikes \citep{Ernst1993a}. This pole will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop. Poles from the Neoproterozoic Chuar Group of southwestern Laurentia (ca. 760 Ma)  as presented in \cite{Eyster2020a}, incorporating data from \cite{Weil2004a}, are also included.

@@

Prior to the termination of the Trans-Hudson orogeny (before 1.8 Ga), paleomagnetic poles need to be considered with respect to the individual Archean provinces. For the Superior province, an additional complexity is that paleomagnetic poles from Siderian to Rhyacian Period (2.50 to 2.05 Ga) dike swarms, as well as deflection of dike trends, support an interpretation that there was substantial Paleoproterozoic rotation of the western Superior province relative to the eastern Superior province across the Kapuskasing Structural Zone \citep{Bates1991a, Evans2010a}. This interpretation is consistent with the hypothesis of \citet{Hoffman1988a} that the Kapuskasing Structural Zone represents major intracratonic uplift related to the Trans-Hudson orogeny. \cite{Evans2010a} propose an Euler rotation of (51\textdegree N, 85\textdegree W, -14\textdegree CCW) to reconstruct western Superior relative to eastern Superior and interpret that the rotation occurred in the time interval of 2.07 to 1.87 Ga.  I follow this interpretation and group the poles into Superior (West) and Superior (East).  Uncertainty remains with respect to whether the ca. 1.88 Ga Molson dikes pole pre-dates or post-dates this rotation \citep{Evans2010a} and thus for the time being should be considered solely in the western Superior province reference frame.

%The poles on this figure should be labeled
\begin{figure*}
\centering
\includegraphics[width=6 in]{../Figures/Fig4_Superior_Slave_reconstructions.pdf}
\caption{\textbf{Paleogeographic reconstructions developed using poles from the Superior, Slave and Rae provinces.} The polarity options that are chosen for the provinces are those that minimize total apparent polar wander path length. This model reconstructs a wide Manikewan ocean that underwent orthogonal closure rather than an alternative possibility of a narrower Manikewan ocean with a pivot-like closure. Paleomagnetic poles are shown colored to match their respective province with these provinces shown in present-day coordinates and labeled in the 0 Ma panel. Poles with ages that are within 25 million years of the given time slice are shown. The relatively well-resolved pole paths from the Superior and Slave provinces (Fig. \ref{fig:Laurentia_poles}) that are utilized for these reconstructions provide strong support for differential plate tectonic motion between 2220 and 1850 Ma.}
\label{fig:Superior_Slave_recons}
\end{figure*}

@@

Following the amalgamation of the Archean provinces in Laurentia by ca. 1.8 Ga, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton \citep{Swanson-Hysell2021b}. It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton. \cite{Halls2015b} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochthon boundary fault. However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of a thick orogenic plateau developed during continent-continent collision (e.g. \citealp{Brown2012a}). Poles with a B-rating are also included in the compilation that come from Greenland, Svalbard and Scotland. These terranes were once part of contiguous Laurentia, but have subsequently been separated through translation and rifting. These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction, and I apply the rotations shown in Table \ref{tab:terrane_rotations}. The Euler pole and rotation is quite well-constrained for Greenland as it is associated with recent opening of Baffin Bay and the Labrador Sea (for which the rotation of \citealp{Roest1989a} is used). The reconstruction of Scotland is associated with the opening of the Atlantic (for which the rotation employed by \citealp{Torsvik2017a} is used) which is well-constrained, but has more uncertainty associated with the Euler pole than that for Greenland. The reconstruction of Svalbard is more challenging and non-unique given a multi-stage tectonic history involving both translation within the Caledonides and subsequent rifting. The preferred Euler pole parameters of \cite{Maloof2006a} are used here for this reconstruction. This Euler rotation is designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland and those of East Svalbard \citep{Maloof2006a, Hoffman2012a} and to reconstruct East Svalbard to be aligned with these correlative sedimentary rocks.

@@

Through the Proterozoic, there are intervals where there are abundant paleomagnetic poles that constrain Laurentia's position and intervals when the record is sparse (shown colored by age in Fig. \ref{fig:Laurentia_poles}). To further visualize the temporal coverage of the poles and to summarize the motion, implied paleolatitudes for an interior point on Laurentia are shown in Figure \ref{fig:Laurentia_paleolatitude}. The ages of the utilized paleomagnetic poles are also shown in comparison to the simplified summary of tectonic events in Figure \ref{fig:tectonic_history}. Both collisional and extensional tectonism can result in the formation of lithologies that can be used to develop paleomagnetic poles either as a result of basin formation, magmatism or both. In addition, intraplate magmatism resulting from plume-related large-igneous provinces (LIPs) can lead to paleomagnetic poles in periods that are otherwise characterized by tectonic quiescence (e.g. the ca. 1267 Ma Mackenzie LIP; Fig. \ref{fig:tectonic_history}). Intracontinental rifts have led to the highest density of poles both in the case of the ca. 1.4 Ga Belt-Purcell Supergroup and the ca. 1.1 Ga Midcontinent Rift (Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). The quality and resolution of the record from the Midcontinent Rift is aided by the voluminous magmatism that occurred in conjunction with basin formation that enables the development of a well-calibrated apparent polar wander path \citep{Swanson-Hysell2019a}. The late Tonian Period also has a number of poles including the Gunbarrel LIP (ca. 780 Ma) and Franklin LIP (ca. 720 Ma), as well as similarly-aged sedimentary rocks from western Laurentia basins \citep{Eyster2020a}. Overall, there is internal consistency among the paleomagnetic poles within intervals for which there is high-resolution coverage. These data result in progressive paths (Fig. \ref{fig:Laurentia_poles}) such as ascending to the ca. 1140 to 1110 Ma apex of the Logan Loop \citep{Robertson1971a}, down the ca. 1110 to 1080 Ma Keweenawan Track \citep{Swanson-Hysell2019a} to the ca. 980 Ma Grenville Loop \citep{McWilliams1975a} prior to a temporal gap before the late Tonian (ca. 775 to 720 Ma) path \citep{Eyster2020a}.

Data from other terranes add resolution to the record. In particular, data from Greenland add 12 poles between 1385 and 1160 Ma when there are only four poles from mainland Laurentia. Given that the rotation between Greenland and mainland Laurentia is well-constrained (Table \ref{tab:terrane_rotations}), once rotated these poles can be used for reconstruction of the entire craton. The reliability of this approach gains credence through the good agreement between the ca. 1633 Ma Melville Bugt diabase dikes pole from Greenland \citep{Halls2011a} and the ca. 1590 Ma Western Channel diabase pole of mainland Laurentia (\citealp{Irving1972a}; Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}). Similarly, there is good agreement between the ca. 1267 Ma Mackenzie dikes pole of Laurentia \citep{Buchan2000a} and coeval poles from Greenland such as the ca. 1275 Ma North Qoroq intrusives \citep{Piper1992a} and Kungnat Ring dike \citep{Piper1977a}. Furthermore, the Greenland poles with ages that fall between the ca. 1237 Ma Sudbury dikes and ca. 1144 Ma lamprophyre dikes pole of mainland Laurentia are consistent with constraints on either side from the mainland while filling in the ascending limb of the path leading up to the apex of 1140 to 1108 Ma poles known as the Logan Loop (Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}).

An exception to this overall agreement between coeval poles from Greenland and mainland Laurentia occurs ca. 1382 Ma. There are poles of this age from Greenland associated with the Zig-Zag Dal basalts and related intrusions \citep{Marcussen1983a, Abrahamsen1987a}. However, these poles are in a distinct location from poles of similar age associated with the Belt-Purcell Supergroup (e.g. the McNamara Formation and Pilcher/Garnet Range and Libby Formations; \citealp{Elston2002a}). Additionally, the older Belt-Purcell Supergroup poles form a more southerly population than time-equivalent poles from elsewhere in Laurentia such as the Mistastin Pluton. There are potential complications associated with the Belt-Purcell Supergroup being exposed within thrust sheets with significant Mesozoic and Cenozoic deformation. However, vertical axis rotations of the Belt region are not able to bring the Belt poles into agreement with those from Laurentia or Greenland nor is translation away from the craton. Another potential complication is that the remanence used for the development of the Belt-Purcell Supergroup poles resides in hematite. As a result, there is the potential for inclination-flatting within the sedimentary rocks from which poles are developed. However, applying a moderate inclination factor of $f=0.6$ also does not bring the poles into congruence with the Zig-Zag Dal basalts. There is the potential that the hematite could be the result of post-depositional oxidation --- the remanence of the Purcell lavas pole is also held by hematite such that it is a chemical remanent magnetization (potentially acquired soon after eruption) rather than being a thermal remanent magnetization held by magnetite \citep{Elston2002a}. However the overall coherency of the pole directions from the Belt-Purcell Supergroup and the presence of reversals as interpreted from antipodal directions has been taken as evidence that the remanence is primary \citep{Elston2002a}. At present, it is unclear which poles are a better representation of Laurentia's position ca. 1400 Ma.

@@

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{../Figures/Fig5_Laurentia_paleolatitude.pdf}
\caption{\textbf{Laurentia paleolatitude through time in data and models.} Top panel: Paleolatitude for the city of Duluth on the southern margin of the Superior province (present-day coordinates of lat=46.79\textdegree N, lon=92.10\textdegree W) implied by paleomagnetic poles from Laurentia and associated terranes. The paleomagnetic poles are compiled in Table 2. Middle panel: Paleolatitude implied by Laurentia poles compared with that implied by published paleogeographic models and the simple Laurentia model used in this chapter for the reconstructions in Figure \ref{fig:Laurentia_reconstructions}. TC2017 refers to \cite{Torsvik2017a} and S-H 2017 refers to \cite{Swanson-Hysell2017a}. Bottom panel: the velocity implied by the continuous paleogeographic models in cm per year for the Duluth reference point on Laurentia.}
\label{fig:Laurentia_paleolatitude}
\end{figure*}

@@

                     Laurentia &                                      Abitibi dikes &     1141$^{+2}_{-2}$ &      A &     279.0 &      48.0 & 215.5 &  48.8 &      14.1 &        55.4 &                                  \cite{Ernst1993a} \\ \hline                       Baltica &                        Salla dike (width 60-100 m) &     1122$^{+5}_{-3}$ &      B &      28.8 &      66.8 & 111.6 &  70.6 &       8.1 &          &                                                 \\ \hline

>>>>

Growth of Laurentia in the Paleoproterozoic also occurred through accretionary orogenesis. In the northwest, this accretion occurred within the Wopmay orogen through ca. 1.88 Ga arc-continent collision that led to the accretion of the Hottah terrane (the Calderian orogeny) and the subsequent emplacement of the Great Bear magmatic zone from ca. 1.88 to 1.84 Ga \citep{Hildebrand2009a}. Coeval with the Trans-Hudson orogeny was accretionary orogenesis on the southern margin of the Superior province (the Penokean orogeny; \citealp{Schulz2007a}) and the southern margin of the North Atlantic province (the Makkovik-Ketilidian orogeny; \citealp{Kerr1996a}). The Penokean orogeny involved the accretion of a microcontinent block (the Marshfield terrane) and arc terranes to the southern margin of the west Superior province ca. 1.86 to 1.82 Ga (Fig. \ref{fig:Laurentia_map}; \citealp{Schulz2007a}). Firm evidence of the end of the Penokean orogeny comes from the ca. 1.78 Ga undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a, Swanson-Hysell2021a}.

@@

Extensive rifting followed by thermal subsidence occurred along the southeastern to eastern Laurentia margin in the time leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean. On the southern part of the eastern margin, a record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig. \ref{fig:Laurentia_map}) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a, Whitmeyer2007a}. The ages of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this history such as in the rift development model of \citet{Burton2010a}. In this model, spatially-restricted rifting occurs ca. 760 to 680 Ma in the region of present-day North Carolina and Virginia. Rifting initiated in the region from present-day New York to Newfoundland ca. 620 to 580 Ma associated with the Central Iapetus Magmatic Province (CIMP; Fig. \ref{fig:Grenville_reconstructions}) and by ca. 580 to 550 Ma rifting extended along the length of Laurentia's eastern margin. The last phase of this rifting has been interpreted to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane ca. 540 Ma \citep{Dickerson1998a, Martin2019a}.

As with other rifted margins, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting and departure of a major craton, as the record that lingers on the craton is similar. Recognizing this ambiguity, it has been proposed that rather than being associated with spatially-restricted or failed rifting, ca. 700 Ma extension (which occured in the southern part of the eastern margin) was associated with breakup and separation of a conjugate continent and that there was subsequent rifting of smaller terranes off the eastern Laurentia margin ca. 600 Ma \citep{Chew2008a, Escayola2011a, Robert2020a}. In these models, the separation of a conjugate continent ca. 700 Ma leads to the formation of an ocean termed the Puncoviscana Ocean by \citet{Escayola2011a} and the Paleo-Iapetus Ocean by \citet{Robert2020a}. The subsequent rifting of terranes off the Laurentia margin ca. 600 Ma would have resulted in the formation of the (Neo-)Iapetus Ocean and the record of thermal subsidence associated with passive margin development on Laurentia \citep{Escayola2011a, Robert2020a}.

@@

In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications (Fig. \ref{fig:Laurentia_poles} and Table 2). The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades to convey the confidence that the community has in these results \citep{Evans2021a}. While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland \citep{Brown2018a}. The `A' rating refers to poles that are judged to be of such high quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. The `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints. In addition, in this chapter I refer to select additional poles that were not given an `A' or `B' classification at the Nordic Workshops as not-rated (`NR'). These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a} in conjunction with the papers in which they were reported. Many additional poles in the database to those that are rated are valuable and should not be dismissed from being considered in paleogeographic reconstructions. However, there are ambiguities associated with many of the poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained, including its age. For example, there are many poles that have been developed from lithologies of the Trans-Hudson orogen (e.g. \citealp{Symons2005a}). However, these poles typically lack field tests, have poor control on paleohorizontal and are hypothesized to suffer from variable overprints as discussed in \cite{Raub2008a} and \cite{DAgrella-Filho2020a}. Given the preponderance of similar directions, three representative poles from the Trans-Hudson orogen are included in the Nordic Workshop compilation (Table 2) as `B' rated poles. However, they should be used with caution given that they suffer from the same deficiencies.

There are similar issues with the rich data sets associated with intrusive and metamorphic lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary. The ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and associated cooling within the Grenville orogen. Cooling ages of deeply exhumed lithologies are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions. As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating with the exception of the `B' rated pole from the ca. 1015 Ma Haliburton intrusions (Table 2). However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within paleogeographic reconstructions of Laurentia (although an alternative view of an allochthonous origin put forward by \citealp{Halls2015a} is discussed below). In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands for which the magnetic mineralogy and associated relative ages of remanence are relatively well-constrained are included (Table 2). Additional not-rated poles, published after the ratings were conferred, that are included in the present compilation are: 1) a pole from the ca. 1780 Ma East Central Minnesota Batholith that confirms that the Slave, Hearne, Rae, and Superior provinces were part of coherent amalgamated Laurentia following the Trans-Hudson orogeny \citep{Swanson-Hysell2021a}; 2) a pole for the ca. 1144 Ma Ontario lamprophyre dikes \citep{Piispa2018a} that strengthens the position of Laurentia at that time and coincides with the position of the poles from the ca. 1140 Ma Abitibi dikes \citep{Ernst1993a}. These poles will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop. Poles from the Neoproterozoic Chuar Group of southwestern Laurentia (ca. 760 Ma)  as presented in \cite{Eyster2020a}, incorporating data from \cite{Weil2004a}, are also included.

@@

Prior to the termination of the Trans-Hudson orogeny (before 1.8 Ga), paleomagnetic poles need to be considered with respect to the individual Archean provinces. For the Superior province, an additional complexity is that paleomagnetic poles from Siderian to Rhyacian Period (2.50 to 2.05 Ga) dike swarms, as well as deflection of dike trends, support an interpretation that there was substantial Paleoproterozoic rotation of the western Superior province relative to the eastern Superior province across the Kapuskasing Structural Zone \citep{Bates1991a, Evans2010a}. This interpretation is consistent with the hypothesis of \citet{Hoffman1988a} that the Kapuskasing Structural Zone represents major intracratonic uplift related to the Trans-Hudson orogeny. \cite{Evans2010a} propose an Euler rotation of (51\textdegree N, 85\textdegree W, -14\textdegree CCW) to reconstruct western Superior relative to eastern Superior and interpret that the rotation occurred in the time interval of 2.07 to 1.87 Ga.  I follow this interpretation and group the poles into Superior (West) and Superior (East).

\begin{figure*}
\centering
\includegraphics[width=5.25 in]{../Figures/Fig4_Superior_Slave_reconstructions.pdf}
\caption{\textbf{Paleogeographic reconstructions developed using poles from the Superior, Slave and Rae provinces.} The polarity options that are chosen for the provinces are those that minimize total apparent polar wander path length. This model follows \cite{Swanson-Hysell2021a} and reconstructs a wide Manikewan ocean that underwent orthogonal closure rather than an alternative possibility of a narrower Manikewan ocean with a pivot-like closure. Paleomagnetic poles are shown colored to match their respective province with these provinces shown in present-day coordinates and labeled in the 0 Ma panel. Poles with ages that are within 20 million years of the given time slice are shown. The relatively well-resolved pole paths from the Superior and Slave provinces (Fig. \ref{fig:Laurentia_poles}) that are utilized for these reconstructions provide strong support for differential plate tectonic motion between 2220 and 1850 Ma.}
\label{fig:Superior_Slave_recons}
\end{figure*}

@@

Following the amalgamation of the Archean provinces in Laurentia by ca. 1.8 Ga, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton. This interpretation of a coherent Laurentia is confirmed paleomagnetically by the consistent positions of ca. 1780 to 1740 Ma poles from the Slave, Hearne/Rae, and Superior Provinces (Fig. \ref{fig:Superior_Slave_recons}; \citealp{Swanson-Hysell2021b}). It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton. \cite{Halls2015b} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochthon boundary fault. However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of a thick orogenic plateau developed during continent-continent collision (e.g. \citealp{Brown2012a}). Poles with a B-rating are also included in the compilation that come from Greenland, Svalbard and Scotland. These terranes were once part of contiguous Laurentia, but have subsequently been separated through translation and rifting. These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction, and I apply the rotations shown in Table \ref{tab:terrane_rotations}. The Euler pole and rotation is quite well-constrained for Greenland as it is associated with recent opening of Baffin Bay and the Labrador Sea (for which the rotation of \citealp{Roest1989a} is used). The reconstruction of Scotland is associated with the opening of the Atlantic (for which the rotation employed by \citealp{Torsvik2017a} is used) which is well-constrained, but has more uncertainty associated with the Euler pole than that for Greenland. The reconstruction of Svalbard is more challenging and non-unique given a multi-stage tectonic history involving both translation within the Caledonides and subsequent rifting. The preferred Euler pole parameters of \cite{Maloof2006a} are used here for this reconstruction. This Euler rotation is designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland and those of East Svalbard \citep{Maloof2006a, Hoffman2012a} and to reconstruct East Svalbard to be aligned with these correlative sedimentary rocks.

@@

Through the Proterozoic, there are intervals where there are abundant paleomagnetic poles that constrain Laurentia's position and intervals when the record is sparse (shown colored by age in Fig. \ref{fig:Laurentia_poles}). To further visualize the temporal coverage of the poles and to summarize the motion, implied paleolatitudes for an interior point on Laurentia are shown in Figure \ref{fig:Laurentia_paleolatitude}. The ages of the utilized paleomagnetic poles are also shown in comparison to the simplified summary of tectonic events in Figure \ref{fig:tectonic_history}. Both collisional and extensional tectonism can result in the formation of lithologies that can be used to develop paleomagnetic poles either as a result of basin formation, magmatism or both. In addition, intraplate magmatism resulting from plume-related large-igneous provinces (LIPs) can lead to paleomagnetic poles in periods that are otherwise characterized by tectonic quiescence (e.g. the ca. 1267 Ma Mackenzie LIP; Fig. \ref{fig:tectonic_history}). Intracontinental rifts have led to the highest density of poles both in the case of the ca. 1.4 Ga Belt-Purcell Supergroup and the ca. 1.1 Ga Midcontinent Rift (Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). The quality and resolution of the record from the Midcontinent Rift is aided by the voluminous magmatism that occurred in conjunction with basin formation that enables the development of a well-calibrated apparent polar wander path \citep{Swanson-Hysell2019a}. The late Tonian Period also has a number of poles including the Gunbarrel LIP (ca. 780 Ma) and Franklin LIP (ca. 720 Ma), as well as similarly-aged sedimentary rocks from western Laurentia basins \citep{Eyster2020a}. Overall, there is internal consistency among the paleomagnetic poles within intervals for which there is high-resolution coverage. These data result in progressive paths (Fig. \ref{fig:Laurentia_poles}) such as ascending to the ca. 1140 to 1108 Ma apex of the Logan Loop \citep{Robertson1971a}, down the ca. 1108 to 1080 Ma Keweenawan Track \citep{Swanson-Hysell2019a} to the ca. 980 Ma Grenville Loop \citep{McWilliams1975a} prior to a temporal gap before the late Tonian (ca. 775 to 720 Ma) path \citep{Eyster2020a}.

Data from other terranes add resolution to the record. In particular, data from Greenland add 12 poles between 1385 and 1160 Ma when there are only four poles from mainland Laurentia. Given that the rotation between Greenland and mainland Laurentia is well-constrained (Table \ref{tab:terrane_rotations}), once rotated these poles can be used for reconstruction of the entire craton. The reliability of this approach gains credence through the good agreement between the ca. 1633 Ma Melville Bugt diabase dikes pole from Greenland \citep{Halls2011a} and the ca. 1590 Ma Western Channel diabase pole of mainland Laurentia (\citealp{Irving1972a}; Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}). Similarly, there is good agreement between the ca. 1267 Ma Mackenzie dikes pole of Laurentia \citep{Buchan2000a} and coeval poles from Greenland such as the ca. 1275 Ma North Qoroq intrusives \citep{Piper1992a} and Kungnat Ring dike \citep{Piper1977a}. Furthermore, the Greenland poles with ages that fall between the ca. 1237 Ma Sudbury dikes and ca. 1144 Ma lamprophyre dikes pole of mainland Laurentia are consistent with older and younger constraints from mainland Laurentia while filling in the ascending limb of the path leading up to the apex of 1140 to 1108 Ma poles known as the Logan Loop (Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}).

An exception to this overall agreement between coeval poles from Greenland and mainland Laurentia occurs ca. 1382 Ma. There are poles of this age from Greenland associated with the Zig-Zag Dal basalts and related intrusions \citep{Marcussen1983a, Abrahamsen1987a}. However, these poles are in a distinct location from poles of similar age associated with the Belt-Purcell Supergroup (e.g. the McNamara Formation and Pilcher/Garnet Range and Libby Formations; \citealp{Elston2002a}). Additionally, the older Belt-Purcell Supergroup poles form a more southerly population than time-equivalent poles from elsewhere in Laurentia such as the Mistastin Pluton. There are potential complications associated with the Belt-Purcell Supergroup being exposed within thrust sheets with significant Mesozoic and Cenozoic deformation. However, vertical axis rotations of the Belt region are not able to bring the Belt poles into agreement with those from Laurentia or Greenland nor is translation away from the craton. Another potential complication is that the remanence used for the development of the Belt-Purcell Supergroup poles resides in hematite. As a result, there is the potential for inclination-flatting within the sedimentary rocks from which poles are developed. However, applying a moderate inclination correction factor of $f=0.6$ also does not bring the poles into congruence with the Zig-Zag Dal basalts. There is the potential that the hematite could be the result of post-depositional oxidation --- the remanence of the Purcell lavas pole is also held by hematite such that it is a chemical remanent magnetization (potentially acquired soon after eruption) rather than being a thermal remanent magnetization held by magnetite \citep{Elston2002a}. However the overall coherency of the pole directions from the Belt-Purcell Supergroup and the presence of geomagnetic reversals as interpreted from antipodal directions has been taken as evidence that the remanence is primary \citep{Elston2002a}. At present, it is unclear which poles are a better representation of Laurentia's position ca. 1400 Ma.

@@

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{../Figures/Fig5_Laurentia_paleolatitude.pdf}
\caption{\textbf{Laurentia paleolatitude through time in data and models.} Top panel: Paleolatitude for the city of Duluth on the southern margin of the Superior province (present-day coordinates of lat=46.79\textdegree N, lon=92.10\textdegree W) implied by paleomagnetic poles from Laurentia and associated terranes. The paleomagnetic poles are compiled in Table 2. Middle panel: Paleolatitude implied by Laurentia poles compared with that implied by published paleogeographic models and the simple Laurentia model used in this chapter for the reconstructions in Figure \ref{fig:Laurentia_reconstructions}. TC2017 refers to \cite{Torsvik2017a} and SHM2017 refers to \cite{Swanson-Hysell2017a}. Bottom panel: the velocity implied by the continuous paleogeographic models in cm per year for the Duluth reference point on Laurentia.}
\label{fig:Laurentia_paleolatitude}
\end{figure*}

@@

                     Laurentia &                                      Abitibi dikes &     1141$^{+2}_{-2}$ &      A &     279.0 &      48.0 & 215.5 &  48.8 &      14.1 &        55.4 &                                  \cite{Ernst1993a} \\ \hline                                            Laurentia &                            Nipigon sills and lavas &     1109$^{+2}_{-2}$ &      A &     270.9 &      49.1 & 217.8 &  47.2 &       4.0 &        56.4 &  Nordic workshop calculation based on data of \cite{Palmer1970a, Robertson1971a, Pesonen1979a, Pesonen1979b, Middleton2004a, Borradaile2006a} \\ \hline
